---
title: "RSF modelis (jaunesnės moterys)"
author: "Domantas Gudas"
date: "2026-03-26"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
---
# Įvadas/kontekstas


# Bibliotekos
```{r, include=FALSE, echo=FALSE}
library(dplyr)
library(ranger)
library(survival)
library(survminer)
library(caret)
library(ggplot2)
library(forcats)
library(timeROC)
library(riskRegression)
library(pec)
library(broom)
library(knitr)
library(kableExtra)
library(tibble)
```

# Duomenų užsikrovimas
```{r, include=FALSE, echo=FALSE}
Final <- read.csv(
  "C:/Users/domgud/Desktop/R/Bakalauras/Final.csv",
  sep = ";"
)

age_median <- median(Final$AMZIUS2016PR, na.rm = TRUE)

df <- Final %>%
  filter(
    AMZIUS2016PR <= age_median,
    ASM_LYTIS == "M"
  )
```

# Duomenų paruošimas
```{r, echo=FALSE}
df_model <- df %>%
  select(
    -ASM_ID,
    -contains("date"),
    -starts_with("CV_death"),
    -starts_with("MACCE"),
    -starts_with("Heart_Failure"),
    -starts_with("AMZIUS2016PR"),
    -starts_with("inpatient"),
    -starts_with("outpatient")
  ) %>%
  mutate(
    death = as.numeric(death),
    death_FU = as.numeric(death_FU)
  )
```

# Mokymo/testavimo aibės
```{r, echo=FALSE}
set.seed(1)
train_idx <- createDataPartition(df_model$death, p = 0.7, list = FALSE)

train_data <- df_model[train_idx, ]
test_data  <- df_model[-train_idx, ]
```

# Modelio apmokymas
```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
set.seed(1)
rsf_model <- ranger(
  Surv(death_FU, death) ~ .,
  data = train_data,
  num.trees = 200,
  mtry = floor(sqrt(ncol(train_data) - 2)),
  splitrule = "logrank",
  importance = "permutation",
  num.threads = max(1, parallel::detectCores() - 1)
)
```

# Konkordacija
```{r, echo=FALSE, warning=FALSE}
set.seed(1)
pred_test <- predict(rsf_model, data = test_data)

test_data$risk_score <- 1 - pred_test$survival[, ncol(pred_test$survival)]

c_index <- survConcordance(
  Surv(death_FU, death) ~ risk_score,
  data = test_data
)

c_val <- round(c_index$concordance, 3)

data.frame(x = paste0("Concordance (C-Index) = ", c_val)) %>%
  kable(format = "html", col.names = NULL) %>%
  kable_styling(position = "center", full_width = FALSE) %>%
  column_spec(1, bold = TRUE, color = "#2E86C1")
```

# Kintamųjų svarba
```{r, echo=FALSE, warning=FALSE, message=FALSE}
perm_imp <- data.frame(
  Variable = names(rsf_model$variable.importance),
  Importance = as.numeric(rsf_model$variable.importance)
) %>%
  arrange(desc(Importance))

perm_imp %>%
  slice(1:20) %>%
  mutate(Variable = fct_reorder(Variable, Importance)) %>%
  ggplot(aes(Variable, Importance)) +
  geom_col(fill = "#4F81BD") +
  coord_flip() +
  theme_minimal(base_size = 14) +
  labs(
    title = "Permutation Importance (Top 20)",
    y = "Δ prediction accuracy",
    x = ""
  )

TOP_N <- 10

top_vars <- perm_imp %>%
  filter(Importance > 0) %>%              
  slice_head(n = TOP_N) %>%
  pull(Variable)
```

# Kaplan-Meier
```{r, echo=FALSE, warning=FALSE, message=FALSE}
top5_vars <- perm_imp$Variable[1:10]

for (var in top5_vars) {
  
  message("Processing: ", var)
  
  if (!var %in% names(test_data)) next
  vals <- na.omit(test_data[[var]])
  if (length(vals) == 0) next
  unique_vals <- unique(vals)
  if (length(unique_vals) < 2) next
  
  is_binary_cat <- is.factor(test_data[[var]]) || 
                   is.character(test_data[[var]]) || 
                   length(unique_vals) <= 2
  
  group_var <- var 
  legend_labels <- NULL 
  
  if (is_binary_cat) {
    test_data[[var]] <- as.factor(test_data[[var]])
    legend_labels <- levels(test_data[[var]])
    
  } else {
    cutp <- tryCatch(
      surv_cutpoint(
        data = test_data,
        time = "death_FU",
        event = "death",
        variables = var,
        minprop = 0.1
      ),
      error = function(e) return(NULL)
    )
    
    if (is.null(cutp)) next
    
    cat_data <- tryCatch(surv_categorize(cutp), error = function(e) return(NULL))
    if (is.null(cat_data)) next
    
    group_var <- paste0(var, "_grp")
    test_data[[group_var]] <- cat_data[[var]]
    legend_labels <- c("Low", "High")
  }
  

  fml <- as.formula(paste("Surv(death_FU, death) ~", group_var))
  
  fit <- survfit(fml, data = test_data)
  
  fit$call$formula <- fml 
  
  pval_res <- survdiff(fml, data = test_data)
  pval_val <- 1 - pchisq(pval_res$chisq, length(pval_res$n) - 1)
  pval_txt <- tryCatch(format.pval(pval_val, digits = 3, eps = 0.001), error=function(e) "NA")

  plot_title <- paste0("Survival by ", var)
  if (!is_binary_cat) {
    cp_val <- signif(cutp$cutpoint$cutpoint, 3)
    plot_title <- paste0(plot_title, " (Cutpoint: ", cp_val, ")")
  }
  plot_title <- paste0(plot_title, "\nLog-rank p = ", pval_txt)

  print(
    ggsurvplot(
      fit,
      data = test_data,
      pval = FALSE,
      risk.table = TRUE,
      conf.int = TRUE,
      palette = if(length(legend_labels) > 2) "jco" else c("#2E86C1", "#E74C3C"),
      title = plot_title,
      legend.labs = legend_labels,
      ggtheme = theme_minimal(),
      tables.theme = theme_minimal()
    )
  )
}
```

# KM grafikas pagal rizikos kvartilius
```{r, echo=FALSE, warning=FALSE, message=FALSE}
test_data <- test_data %>%
  mutate(
    risk_group = cut(
      risk_score,
      breaks = quantile(risk_score, probs = seq(0, 1, 0.25), na.rm = TRUE),
      include.lowest = TRUE,
      labels = c("Q1 (Low)", "Q2", "Q3", "Q4 (High)")
    )
  )

fit_risk <- survfit(Surv(death_FU, death) ~ risk_group, data = test_data)

fit_risk$call$formula <- Surv(death_FU, death) ~ risk_group

pval_risk <- survdiff(
  Surv(death_FU, death) ~ risk_group,
  data = test_data
)

pval_txt <- signif(
  1 - pchisq(pval_risk$chisq, length(pval_risk$n) - 1),
  3
)

ggsurvplot(
  fit_risk,
  data = test_data,
  pval = FALSE,
  risk.table = TRUE,
  palette = c("#1B9E77", "#D95F02", "#7570B3", "#E7298A"),
  title = paste(
    "Survival by Predicted Risk Quartiles (log-rank p =",
    pval_txt, ")"
  ),
  ggtheme = theme_minimal(),
  tables.theme = theme_minimal()
)

```

# Nuo laiko priklausantys AUC
```{r, echo=FALSE, warning=FALSE, message=FALSE}
times_eval <- c(365, 730, 1095, 1459)

roc_res <- timeROC(
  T = test_data$death_FU,
  delta = test_data$death,
  marker = test_data$risk_score,
  cause = 1,
  weighting = "marginal",
  times = times_eval,
  iid = FALSE
)

plot_df <- data.frame()

for (i in seq_along(times_eval)) {
  auc_val <- round(roc_res$AUC[[i]], 3)
  label <- paste0(round(times_eval[i]/365, 1), " Years (AUC = ", auc_val, ")")
  
  temp_df <- data.frame(
    FP = roc_res$FP[, i],
    TP = roc_res$TP[, i],
    Timepoint = label, 
    AUC = auc_val   
  )
  plot_df <- rbind(plot_df, temp_df)
}

auc_table <- data.frame(
  `Follow-up Time` = c("1 Year", "2 Years", "3 Years", "4 Years"),
  `Days` = times_eval,
  `AUC Score` = round(roc_res$AUC, 3)
)

auc_table %>%
  kable(
    format = "html", 
    col.names = c("Follow-up Interval", "Days", "Area Under the Curve (AUC)"),
    caption = "Performance Summary: Time-Dependent Discrimination Power",
    align = "lcc"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  row_spec(0, bold = TRUE, color = "white", background = "#2E86C1") %>% 
  column_spec(3, bold = TRUE, color = "#2E86C1") %>% 
  footnote(general = "AUC values represent the model's ability to distinguish between survivors and non-survivors at specific time points.")

ggplot(plot_df, aes(x = FP, y = TP, color = Timepoint)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray60") +
  geom_line(size = 1.2) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Time-Dependent ROC Curves",
    subtitle = "Prediction performance at 1, 2, 3, and 4 years",
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)",
    color = "Timepoint"
  ) +
  theme_minimal() +
  theme(
    legend.position = c(0.7, 0.2), 
    legend.background = element_rect(fill = "white", color = "gray90"),
    plot.title = element_text(face = "bold", size = 14),
    panel.grid.minor = element_blank()
  )
```

# Kalibracijos grafikai (1-4 metai)
```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=8}
library(riskRegression)
library(ggplot2)
library(dplyr)
library(scales)

times_eval <- c(365, 730, 1095, 1459)

cal_all <- Score(
  object = list("RSF" = rsf_model),
  formula = Surv(death_FU, death) ~ 1,
  data = test_data,
  times = times_eval,
  metrics = "Brier",
  plots = "Calibration",
  cens.method = "ipcw",
  keep.matrix = TRUE
)


cal_df <- cal_all$Calibration$plotframe %>%
  filter(model == "RSF", times %in% times_eval) %>%
  mutate(time_label = paste0(round(times / 365, 1), " Year(s)"))

cal_plot_df <- cal_df %>%
  mutate(risk_group = ntile(risk, 10)) %>%
  group_by(times, time_label, risk_group) %>%
  summarise(
    pred_risk = mean(risk, na.rm = TRUE),
    obs_risk = sum(riskRegression_status * Wt) / sum(Wt),
    .groups = "drop"
  )

ggplot(cal_plot_df, aes(x = pred_risk, y = obs_risk)) +
  geom_point(color = "#E74C3C", size = 3) +
  geom_line(color = "#2E86C1", size = 1.2) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50") +
  facet_wrap(~time_label) +
  scale_x_continuous(labels = percent_format(accuracy = 1), limits = c(0, NA)) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
  labs(
    title = "Calibration Plot for RSF Model",
    subtitle = "Red points: mean risk per decile | Blue line: smoothed calibration",
    x = "Predicted Risk",
    y = "Observed Risk"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.background = element_rect(fill = "#F8F9F9", color = "gray90"),
    strip.text = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank(),
    aspect.ratio = 1
  )

```

# Standartinis Cox model
```{r, echo=FALSE, warning=FALSE, message=FALSE}
cox_vars <- intersect(top_vars, names(train_data))

train_data[cox_vars] <- lapply(train_data[cox_vars], function(x) {
  if (is.factor(x)) as.numeric(as.character(x))
  else if (is.logical(x)) as.numeric(x)
  else x
})


cox_formula <- as.formula(
  paste("Surv(death_FU, death) ~", paste(cox_vars, collapse = " + "))
)

cox_model <- coxph(cox_formula, data = train_data)

cox_results <- tidy(cox_model, exponentiate = TRUE, conf.int = TRUE) %>%
  select(
    Variable = term,
    HR = estimate,
    `95% CI Lower` = conf.low,
    `95% CI Upper` = conf.high,
    `p-value` = p.value
  ) %>%
  mutate(
    across(c(HR, `95% CI Lower`, `95% CI Upper`), ~ round(., 2)),
    `p-value` = case_when(
      `p-value` < 0.001 ~ "< 0.001 ***",
      `p-value` < 0.01  ~ paste0(round(`p-value`, 3), " **"),
      `p-value` < 0.05  ~ paste0(round(`p-value`, 3), " *"),
      TRUE ~ as.character(round(`p-value`, 3))
    )
  )

cox_results %>%
  kable(
    format = "html", 
    caption = "Multivariable Cox Proportional Hazards Regression Results",
    align = "lcccc"
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  column_spec(1, bold = TRUE) %>%
  column_spec(2, bold = TRUE, color = "white", background = "#2E86C1") %>% 
  add_header_above(c(" " = 1, "Effect Size" = 1, "Confidence Interval" = 2, "Significance" = 1)) %>%
  footnote(general = "* p < 0.05, ** p < 0.01, *** p < 0.001. HR = Hazard Ratio.")
```

# Cox modelio prielaidos
```{r, echo=FALSE, warning=FALSE, message=FALSE}
cox_zph <- cox.zph(cox_model)

zph_table <- as.data.frame(cox_zph$table) %>%
  rownames_to_column("Variable") %>%
  mutate(
    Chi_Square = round(chisq, 3),
    p_val = as.numeric(p),
    `p-value` = ifelse(p < 0.001, "< 0.001", as.character(round(p, 3))),
    Status = ifelse(
      is.na(p_val), 
      "—",
      ifelse(p_val < 0.05, "✘ Fail (Violated)", "✔ Pass (Proportional)")
    )
  ) %>%
  select(Variable, Chi_Square, `p-value`, Status)

zph_table %>%
  kable(
    format = "html",
    caption = "Schoenfeld Residuals Test for Proportional Hazards Assumption"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  column_spec(
    4,
    color = ifelse(zph_table$Status == "✔ Pass (Proportional)", "green", "red"),
    bold = TRUE
  )

zph_logic <- as.data.frame(cox_zph$table) %>%
  rownames_to_column("Variable") %>%
  filter(Variable != "GLOBAL") %>%
  mutate(p_val = as.numeric(p))

tv_vars <- zph_logic %>%
  filter(p_val < 0.05) %>%
  pull(Variable)

const_vars <- zph_logic %>%
  filter(p_val >= 0.05) %>%
  pull(Variable)

```

# Laiko priklausomi kintamieji (Time-Varying Effects)
```{r, echo=FALSE, warning=FALSE, message=FALSE}
cox_tv_formula <- as.formula(
  paste(
    "Surv(death_FU, death) ~",
    paste(const_vars, collapse = " + "),
    if (length(tv_vars) > 0)
      paste("+", paste(paste0("tt(", tv_vars, ")"), collapse = " + "))
  )
)


cox_tv <- coxph(
  cox_tv_formula,
  data = train_data,
  tt = function(x, t, ...) x * log(t + 1)
)


tv_results <- tidy(cox_tv, conf.int = TRUE) %>%
  mutate(
    Effect_Type = ifelse(grepl("^tt\\(", term),
                         "Time-varying (log(t+1))",
                         "Constant"),
    Variable = gsub("tt\\(|\\)", "", term),
    Estimate = round(estimate, 3),
    `Std. Error` = round(std.error, 3),
    `z-statistic` = round(statistic, 2),
    `p-value` = case_when(
      p.value < 0.001 ~ "< 0.001 ***",
      p.value < 0.01  ~ "< 0.01 **",
      p.value < 0.05  ~ "< 0.05 *",
      TRUE ~ as.character(round(p.value, 3))
    )
  ) %>%
  select(Variable, Effect_Type, Estimate, `Std. Error`, `z-statistic`, `p-value`)


tv_results <- tidy(cox_tv, conf.int = TRUE) %>%
  mutate(
    Effect_Type = ifelse(grepl("^tt\\(", term), 
                         "Time-varying (log(t+1))", 
                         "Constant"),
    Variable = gsub("tt\\(|\\)", "", term),
    
    Display_Value = ifelse(Effect_Type == "Constant", 
                           round(exp(estimate), 2), 
                           round(estimate, 3)),
    
    Conf_Low = ifelse(Effect_Type == "Constant", round(exp(conf.low), 2), round(conf.low, 3)),
    Conf_High = ifelse(Effect_Type == "Constant", round(exp(conf.high), 2), round(conf.high, 3)),
    
    `p-value` = case_when(
      p.value < 0.001 ~ "< 0.001 ***",
      p.value < 0.01  ~ "< 0.01 **",
      p.value < 0.05  ~ "< 0.05 *",
      TRUE ~ as.character(round(p.value, 3))
    )
  ) %>%
  mutate(Result = ifelse(Effect_Type == "Constant",
                         paste0(Display_Value, " (", Conf_Low, "-", Conf_High, ")"),
                         as.character(Display_Value))) %>%
  select(Variable, Effect_Type, Result, `p-value`)

tv_results %>%
  kable(
    format = "html",
    caption = "Final Clinical Model: Constant Hazard Ratios & Time-Varying Slopes",
    col.names = c("Variable", "Effect Type", "HR (95% CI) or Slope", "Significance"),
    align = "llcc"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = FALSE) %>%
  footnote(
    general = "For 'Constant' variables, the value represents the Hazard Ratio. For 'Time-varying' variables, the value represents the interaction slope with log(time); positive values indicate risk increases over time."
  )
```

# AFT
```{r, echo=FALSE, warning=FALSE, message=FALSE}
aft_formula <- as.formula(
  paste("Surv(death_FU, death) ~", paste(cox_vars, collapse = " + "))
)

fit_weibull   <- survreg(aft_formula, data = train_data, dist = "weibull")
fit_lognormal <- survreg(aft_formula, data = train_data, dist = "lognormal")

```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
weibull_exp   <- exp(coef(fit_weibull))
lognormal_exp <- exp(coef(fit_lognormal))

exp_table <- data.frame(
  Variable  = names(weibull_exp)[-1],
  Weibull   = round(weibull_exp[-1], 3),
  Lognormal = round(lognormal_exp[-1], 3)
)


kable(exp_table, caption = "Exponentiated AFT Model Coefficients (Weibull vs Lognormal)")
```


